PACKER: An Exemplar Model of Category Generation

### 2017-01-23 Manuscripts/cogsci-2017/paper.tex
intro reorganize

In one of the earliest explorations on the creative use of conceptual knowledge, Ward \& colleagues \citep{marsh1999inadvertent,smith1993constraining,ward2002role,ward1994structured} provided clear evidence of the role of prior knowledge in concept generation. In a canonical study, \citet{ward1994structured} asked participants to draw and describe novel species of plants and animals that might exist on other planets. Generation was strongly constrained by prior knowledge of Earth plants and animals: people generated alien species using the same structural forms found on Earth (e.g., eyes, legs, wings), exemplars tended to obey the same feature correlations observed on Earth (e.g., feathers tended to co-occur with wings), and exemplars drawn from the same species were less variable than animals drawn from different species (as is the case on Earth).

% early formal models, intro copy and tweak. [joe: did you ever get those papers?]

In a more recent report, \citet{jern2013probabilistic} developed a hierarchical model capable of explaining the influence of prior knowledge observed by \citet{ward1994structured}. In their model, categories are represented as distributions in multidimensional space, where known exemplars are assumed to have been generated from the underlying category distribution. Category distributions are in turn thought to be generated by an underlying domain distribution, representing common distributional characteristics of known categories (e.g., feature variance, feature correlations). When asked to generate exemplars from a novel category, people are thought to sample a new category from the domain distribution, and then sample exemplars from the category distribution. Thus, the generated exemplars tend to obey the same distributional properties of known categories.

\citet{jern2013probabilistic} reported several experiments in support of their model's predictions. In their paradigm, participants are first exposed to members from a set of known categories within a $k$-dimensional feature space (in their experiments, $k=3$). After observing the assigned examples, participants are asked to generate exemplars from a new category within the domain. Participants presented with a set of scales scales that can be used to adjust the feature values of the generated stimulus, and are given unlimited time to create each example. As in the classic \cite{ward1994structured} experiment, \citet{jern2013probabilistic} found that people generated categories possessing the same feature correlations as the known categories in the domain.

The work presented by Ward \& Colleagues, as well as that of \citet{jern2013probabilistic}, undoubtedly captures a core principle of the creative use of conceptual knowledge: category generation is informed by the distributional characteristics of known concepts in the target domain. However, this work fails to describe the inverse: how do generated categories \textit{differ} from known categories? In this paper, we introduce a novel exemplar-based approach to category generation, the PACKER model (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}), which creates new categories by balancing two constraints: (1) new categories should be different from known categories (minimizing between-class similarity), and (2) new categories should be internally coherent (maximizing within-class similarity). We then report a behavioral experiment testing the predictions of the PACKER model, and report formal simulations comparing PACKER to the model proposed by \cite{jern2013probabilistic}.

@@

\begin{figure*}
    \begin{center}
    \inputpgf{figs/}{example-prob-spaces.pgf}
    \caption{PACKER generation of a category `B' example, following exposure to one member of category `A' and one member of category `B'. Areas in which generation is not likely are shaded white; high probability areas are shaded blue. \textit{Left}: Predictions given $\{\phi = -1$, $\gamma = 0\}$ (contrast influence only). \textit{Center}: Predictions given $\{\phi = 0$, $\gamma = 1\}$ (target influence only).  \textit{Right}: Predictions given $\{\phi = -1$, $\gamma = 1\}$ (both constraints considered).  }
    \label{fig:example-prob-spaces}
    \end{center}
\end{figure*}

@@

% Note something about euclidean distance for integral dims?  Or is the field over that?

@@

The PACKER model described above explains exemplar and category generation under the traditional exemplar-similarity approach that has guided much of the field's theoretical and empirical work for over three decades. The model proposes that people generate categories by maximizing within-category similarity and minimizing between-category similarity. See Figure \ref{fig:example-prob-spaces} for a depiction of the model's behavior.

In the sections below, we report a behavioral experiment testing a key difference between PACKER and previous accounts: unlike the model proposed by \citeauthor{jern2013probabilistic}, the distributional properties of known categories affects performance in PACKER only insofar as it constrains inter-exemplar similarities. Instead, the \textit{location} of known categories in the domain is crucial because it constrains remaining possible locations for new categories. We then report formal simulations evaluating the fit of the PACKER and \citet{jern2013probabilistic} models to our data.

@@

The behavioral experiment described below was designed to test whether the \textit{location} of contrast categories (as opposed to their structure), influences generation. As such, our experiment manipulates only the location of the known categories within the domain (the structure of the categories is held constant).

@@

The experiment follows the paradigm set by \citet{jern2013probabilistic}: first, participants are exposed to members of a known category (`Alpha', or `A'), and are then asked to generate exemplars belonging to a new category (`Beta', or `B'). In both conditions, members of the Alpha category are tightly clustered, with equal variance on both features and no correlation between features. Our manipulation is fairly `weak' (see Figure \ref{fig:middle-bottom-conditions}): the conditions have only a slight difference in the Y-Axis position of the Alpha category. In the `Middle' condition the Alpha category is placed in the center of the space, in the `Bottom' condition the Alpha category is placed in the bottom-center of the space.

Because the distributional characteristics of the domain do not differ between the Middle and Bottom conditions, the \citet{jern2013probabilistic} model predicts that there should be no difference in the categories generated within each domain. However, the PACKER model proposes that exemplars are generated such that each category has strong within-class similarity and weak between-class similarity. By consequence, the upper areas of the stimulus space should be used for generation more frequently in the Bottom condition compared to Middle, as stimuli in this area possess weaker similarity to members of the Alpha category. Conversely, the bottom areas of the space should be less frequently used for generation in the Bottom condition compared to Middle, as stimuli in this area possess greater similarity to the Alpha category. 
% there has GOT to be a more concise way of expressing those final two sentences...

>>>>

In one of the earliest explorations on the creative use of conceptual knowledge, Ward \& colleagues \citep{marsh1999inadvertent,smith1993constraining,ward2002role,ward1994structured,ward1995s} provided clear evidence of the role of prior knowledge in concept generation. In a canonical study, \citet{ward1994structured} asked participants to draw and describe novel species of plants and animals that might exist on other planets. Generation was strongly constrained by prior knowledge of Earth plants and animals: people generated alien species using the same structural forms found on Earth (e.g., eyes, legs, wings), exemplars tended to obey the same feature correlations observed on Earth (e.g., feathers tended to co-occur with wings), and exemplars drawn from the same species were less variable than animals drawn from different species (as is the case on Earth).

An early account of these these results proposed that people complete the generation task by retrieving memories of known Earth plants and animals, and slightly changing them to create something new \citep{ward2002role,ward1995s}.  This account, which we will refer to as the \textit{copy-and-tweak} account can be casted within a formal exemplar approach \citep[see][]{jern2013probabilistic}. However, due to the nature of the response formats used (e.g., drawing and describing examples), and due the prior knowledge assumed (e.g., knowledge of plants and animals on Earth), these experiments are not suitable for formal modeling.

In more recent work, \citet{jern2013probabilistic} reported several experiments on exemplar and category generation within an artificial category learning paradigm (and are thus suitable for formal modeling). In their paradigm, participants are first exposed to members from a set of known categories within a reduced feature space (e.g., size, hue, saturation). After observing the assigned examples, participants are asked to generate exemplars from a new category within the domain. Participants provided with a set of scales scales that can be used to adjust the feature values of the generated stimulus (one scale per feature), and are given unlimited time to create each example. As in the classic \cite{ward1994structured} experiment, \citet{jern2013probabilistic} found that people generated categories possessing the same feature correlations as the known categories in the domain.

\citet{jern2013probabilistic} developed a hierarchical model of generation in order to explain their results. In their model, categories are represented as probability distributions in a multidimensional space (with each stimulus feature corresponding to a dimension of the space). Observed exemplars (represented as points in this space) are viewed as samples from an underlying category distribution (e.g., a multivariate normal distribution). These category distributions are in turn viewed as samples from an underlying domain distribution, which represents the common distributional characteristics among known categories (e.g., feature variance, feature correlations). When asked to generate a novel class, people are thought to sample a category distribution from the domain distribution, and then sample exemplars from the category distribution. Thus, exemplars generated into novel classes tend to obey the same distributional properties of known categories.

The work presented by Ward \& Colleagues, as well as that of \citet{jern2013probabilistic}, undoubtedly captures a core principle of the creative use of conceptual knowledge: category generation is informed by prior knowledge of concepts in the target domain. The two models described above (copy-and-tweak, hierarchical sampling) are conceptually similar in that they assume learners \textit{copy} something about existing categories: individuals either directly copy-and-tweak stored observations, or copy the distributional structure of entire categories.

\begin{figure*}
    \begin{center}
    \inputpgf{figs/}{example-prob-spaces.pgf}
    \caption{PACKER generation of a category `B' example, following exposure to one member of category `A' and one member of category `B'. \textit{Left}: Predictions given $\{\phi = -1$, $\gamma = 0\}$ (contrast influence only). \textit{Center}: Predictions given $\{\phi = 0$, $\gamma = 1\}$ (target influence only).  \textit{Right}: Predictions given $\{\phi = -1$, $\gamma = 1\}$ (both constraints considered).  }
    \label{fig:example-prob-spaces}
    \end{center}
\end{figure*}

In this paper, we introduce a novel exemplar-based approach to category generation, the PACKER model (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}), which creates new categories by balancing two constraints: (1) new categories should be different from known categories (minimizing between-class similarity), and (2) new categories should be internally coherent (maximizing within-class similarity). PACKER is a significant departure from the accounts of generation reviewed above in that it instantiates assumptions about how exemplars in novel classes should \textit{differ} from existing categories (rather than how they should align to them). Likewise, the PACKER model is of interest because it follows directly from the well-studied exemplar-similarity framework that has led the field for over three decades.

In the sections below, we formally describe the PACKER model. We then report a behavioral experiment testing the model's predictions, and report formal simulations comparing PACKER to the copy-and-tweak and hierarchical sampling models.

@@

@@

% Note something about minkowski metric? Or can we assume people know that.

@@

The PACKER model described above explains exemplar and category generation under a traditional exemplar-similarity approach. The model proposes that people generate categories by maximizing within-category similarity and minimizing between-category similarity.

PACKER is unique as a model of generation because it specifies precisely how generated classes should \textit{differ} from existing ones -- the copy-and-tweak and hierarchical sampling accounts explain why generated classes share many properties with known classes, but not how they should be distinct. Likewise, PACKER is unique in that the distributional properties of known categories affect generation only insofar as they constrain inter-exemplar similarities. Instead, the \textit{location} of known categories in the domain is crucial because it constrains remaining possible locations for new categories. In the sections below, we report a behavioral experiment testing these predictions. We then report formal simulations with PACKER, a copy-and-tweak model, and the \citet{jern2013probabilistic} hierarchical sampling model.

@@

The behavioral experiment described below was designed to test whether the \textit{location} of contrast categories (as opposed to their structure), influences generation. The experiment follows the paradigm developed by \citet{jern2013probabilistic}: first, participants are exposed to members of a known category (`Alpha', or `A'), and are then asked to generate exemplars belonging to a new category (`Beta', or `B').

We developed two Alpha categories (see Figure \ref{fig:middle-bottom-conditions}), instantiating the two conditions of the experiment. In both conditions, members of the Alpha category are tightly clustered, with equal variance on both features and no correlation between features. Our manipulation is fairly `weak': the conditions have only a slight difference in the Y-Axis position of the Alpha category. In the `Middle' condition the Alpha category is placed in the center of the space, in the `Bottom' condition the Alpha category is placed in the bottom-center of the space.

@@

Although our manipulation is minimal, the PACKER model is capable of predicting strong between-condition differences. The PACKER model proposes that the nature of the space not occupied by the Alpha category determines where members of the Beta category are likely to be generated. Thus, the lower areas of the stimulus space should be less frequently used for generation in the Bottom condition compared to the Middle (as these areas possess greater similarity to the Bottom Alpha category). Conversely, the upper areas of the stimulus space should be used for generation more frequently in the Bottom condition compared to Middle.

More generally, the PACKER model proposes that the probability a stimulus $y$ will be generated is a function of its similarity to contrast categories \textit{and} to members of the target category. Two more general predictions (not specific to either condition) follow from this proposal: (1) the location of beta examples should be positively related to distance from the Alpha category, (2) Beta examples should be more similar to one another than they are to members of the Alpha category.

% THIS IS WHERE NOLAN STOPPED EDITING AFTER THE INTRO REFORMULATION
